---
title: "Modelling Cryptocurrency exchange rates using ARMA Models"
subtitle: "Applied Macroeconometrics Assignment I"
author: "Rishabh Patil | 2021A7PS0464H"
layout: page
output:
  pdf_document: 
    includes:
      in_header: "wrap-code.tex"
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_tex: yes
    highlight: espresso
    df_print: kable
    fig_crop: no
  html_document: 
    toc: yes
    highlight: textmate
    theme: simplex
    fig_width: 10
    fig_height: 7
    fig_caption: yes
  html_notebook: 
    toc: yes
    highlight: espresso
    theme: united
    fig_caption: yes
    number_sections: yes
editor_options:
  chunk_output_type: inline
---

```{r}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)

```

```{r}
#install.packages("quantmod")
```

```{r warning=FALSE}
library(quantmod)
```

# Description of Dataset

For our study, we'd be analyzing the exchange rates of the cryptocurrency Bitcoin, more precisely the weekly returns.

**The variable of concern here is the price(in USD) of Bitcoin (returns on it).**

Since there is a general notion of cryptocurrencies having a 4 year cycle, we'd be looking at 8-10 year data. And to make our analysis feasible, instead of a daily chart, we would be analyzing a weekly rolling average for our time period of 10 years.

## The data

```{r}
btc_df <- getSymbols("BTC-USD", src = "yahoo", from = Sys.Date() - 365 * 10, to = Sys.Date(), auto.assign = FALSE)
print(head(btc_df))
```

### Candlestick Chart (daily pricing)

```{r fig.height=10, fig.width=14}
chartSeries(btc_df, name="BTC-USD", subset="last 120 months")
```

### Line Graph(daily pricing)

```{r fig.height=10, fig.width=14}
library(ggplot2)

ggplot(data=btc_df,aes(y=btc_df$`BTC-USD.Close`,x=index(btc_df)),group=1)+
  geom_line()

```

### Clustering Month-wise avg price:

```{r}
#install.packages("xts")
```

```{r}
library(xts)
btc_week_df<- apply.weekly(btc_df,FUN = mean)
head(btc_week_df)
```

### Line Graph

```{r fig.height=12, fig.width=18}
ggplot(data=btc_week_df,aes(y=btc_week_df$`BTC-USD.Close`,x=index(btc_week_df)),group=1)+
  geom_line()
```

### Candlestick Graph

```{r fig.height=10, fig.width=14}
chartSeries(btc_week_df)
```

### Returns

For plotting the returns, we use this formula:

$$
R = \frac{S(t+1)-S(t)}{S(t)} \text{ or plotting the log return: } RLt = ln\biggl(\frac{S(t+1)}{S(t)}\biggr)
$$

```{r}
daily_returns <- log(btc_df/lag(btc_df,1))
ggplot(data=daily_returns,aes(y=daily_returns$`BTC-USD.Close`,x=index(daily_returns)),group=1)+
  geom_line()
```

```{r}
weekly_return <- log(btc_week_df$`BTC-USD.Close`/lag(btc_week_df$`BTC-USD.Close`,1))
ggplot(data=weekly_return,aes(y=weekly_return$`BTC-USD.Close`,x=index(weekly_return)),group=1)+
  geom_line()
```

This is the basis of our analysis

## Visual Interpretation

From the above log returns chart we can see that the long run expectation for weekly return is 0

$$
\mathop{\mathbb{E}}[y_t]=0
$$

```{r}
ggplot(weekly_return, aes(weekly_return$`BTC-USD.Close`)) +
  geom_histogram()
```

The data looks normally distributed with slight outliers.

**looking for seasonality:**

```{r}
yearly_return<-yearlyReturn(btc_df)
ggplot(data=yearly_return,aes(y=yearly_return$yearly.returns,x=index(yearly_return)),group=1)+
  geom_line()
```

We do see that there is a 4-year cycle apparent from the yearly returns, that is, the returns rise up for a year, then reach peak very steeply the next and a steep decline sets in ending the cycle.

**But apart from the yearly return analysis the weekly returns appear stationery.**

The objective of this study is to model the weekly returns with appropriate ARMA model(s) and predict the trend (forecast it).

### Estimation and Holdback dataset

**Estimation:** from 2014-09-21 to 2022-09-21 - *train data*

**Holdback:** from 2022-09-28 to 2024-02-23 - *test data*

(\~75/25 split) (split post cycle)

```{r}
btc_week_est <- btc_week_df[index(btc_week_df) >= as.Date("2014-09-21") & 
                                index(btc_week_df) <= as.Date("2022-09-21"), ]

btc_week_hb <- btc_week_df[index(btc_week_df) >= as.Date("2022-09-22") & 
                              index(btc_week_df) <= as.Date("2024-02-23"), ]
```

```{r}
chartSeries(btc_week_est)
```

```{r}
chartSeries(btc_week_hb)
```

```{r}
weekly_return_est <- weekly_return[index(weekly_return) >= as.Date("2014-09-21") & 
                                index(weekly_return) <= as.Date("2022-09-21"), ]
weekly_return_hb <- weekly_return[index(weekly_return) >= as.Date("2022-09-22") & 
                                index(weekly_return) <= as.Date("2024-02-23"), ]
```

```{r}
ggplot(data=weekly_return_est,aes(y=weekly_return_est$`BTC-USD.Close`,x=index(weekly_return_est)),group=1)+
  geom_line()
```

```{r}
ggplot(data=weekly_return_hb,aes(y=weekly_return_hb$`BTC-USD.Close`,x=index(weekly_return_hb)),group=1)+
  geom_line()
```

# Model fitting and model selection

## Analyzing ACF and PACF

```{r}
acf(weekly_return_est[2:length(weekly_return_est)])
```

**We see that the *ACF is smoothly decaying* which means its more *likely to be an AR process***

```{r}
pacf(weekly_return_est[2:length(weekly_return_est)])
```

There is no continuous decay in PACF, the value abruptly falls after first lag. so we can assume it to be a AR process. **Note:** *The 19th lag is also significant, suggesting some seasonality. But we will ignore that for now.*

Since one lag is significant, AR(1) model can be tested. We will also test ARMA(1,1) and ARMA(1,2) and ARMA(2,1)

## Testing ARMA models

### AR(1)

```{r}
ar1<-arima(weekly_return_est$`BTC-USD.Close`,order=c(1,0,0))
summary(ar1)
```

#### Coefficients Test

The coefficients have some standard error but can be considered insignificant. Statistically testing the significance:

```{r}
library(lmtest)
coeftest(ar1)
```

The coefficients are statistically significant. (The Intercept has a lower significance threshold, but can be fitted in for the model.)

#### Residual Check

```{r}
library('forecast')
checkresiduals(ar1)
```

The residuals seem to be normally distributed and have a mean of 0. The ACF has an outlier at 19th lag(as discussed earlier maybe due to seasonality).

#### Stationarity, Invertibility and Causality

Since AR(1), we only need to check for stationarity and causality (only pertaining to $\phi(z)$)

$$
\phi(z)y_t = \theta(z)u_t
$$

```{r}
autoplot(ar1)
```

roots within the circle, stationary and non-causal.

#### Mean of the model

$$
\mathop{\mathbb{E}}[y_t \sim ARMA(p,q)]=\mathop{\mathbb{E}}[y_t \sim AR(p)] = \frac{a_0}{1-\sum_{i=1}^pa_i}
$$

```{r}
ar1$coef
```

$$
= -0.009/0.2808 \approx 0.3205
$$

### ARMA(2,1)

```{r}
arma21<-arima(weekly_return_est$`BTC-USD.Close`,order=c(2,0,1))
arma21
```

The coefficients have very high standard error

#### Coefficients Test

```{r}
coeftest(arma21)
```

The coefficients are not significant.

#### Residual Check

```{r}
checkresiduals(arma21)

```

#### Stationarity, Invertibility and Causality

```{r}
autoplot(arma21)
```

non-causal stationary,non-invertible.

#### Mean

=

```         
0.0092889/(1-(0.1362553+0.0495093))
= 0.01140812595
```

### ARMA(1,2)

```{r}
arma12<-arima(weekly_return_est$`BTC-USD.Close`,order=c(1,0,2))
arma12
```

#### Coefficients Test

```{r}
coeftest(arma12)
```

Coefficients are not significant.

#### Residual Check

```{r}
checkresiduals(arma12)
```

#### Stationarity, Invertibility and Causality

```{r}
autoplot(arma12)
```

non-Invertible, stationary and not causal

#### Mean

```         
= 0.0092838/(1-0.3042082)
= 0.01334278443
```

### MA(2)

```{r}
ma2 <-arima(weekly_return_est$`BTC-USD.Close`,order=c(0,0,2))
summary(ma2)
```

Moderately High Standard Errors

#### Coefficients Test

```{r}
coeftest(ma2)
```

The ma1 coefficient appears to be significant that too with a narrower CI, and the ma2 and intercept appear to be significant with loose constraints.

#### Residual Check

```{r}
checkresiduals(ma2)
```

Even the residuals appear to be normally distributed at zero.

#### Stationarity, Causality and Invertibility

```{r}
autoplot(ma2)
```

There is no invertibility and stationarity is also met. (no causality as well)

#### Mean

= 0

### Ljung-Box test for AR(1) MA(2) and ARMA(1,2)

```{r}
print(Box.test(ar1$resid, type="Ljung-Box", lag=20))
print(Box.test(ma2$resid, type="Ljung-Box", lag=20))
print(Box.test(arma12$resid, type="Ljung-Box", lag=20))
```

```{r}
d1<- checkresiduals(ar1)$statistic
d2<- checkresiduals(ma2)$statistic
d3<- checkresiduals(arma12)$statistic
print(d1)
print(d2)
print(d3)
```

#### AIC

```{r}
print(AIC(ar1))
print(AIC(ma2))
print(AIC(arma12))
```

+-------------------+-----------------+----------------+----------------+
| Col1              | ARMA(1,2)       | AR(1)          | MA(2)          |
+===================+=================+================+================+
| AR Coeff 1        | 0.3042(1.0178)  | 0.2808(0.0469) | \-             |
+-------------------+-----------------+----------------+----------------+
| AR Coeff 2        | \-              | \-             | \-             |
+-------------------+-----------------+----------------+----------------+
| MA Coeff 1        | -0.0259(1.0238) | \-             | 0.2777(0.0487) |
+-------------------+-----------------+----------------+----------------+
| MA Coeff 2        | -0.0016(0.2902) | \-             | 0.0806(0.0474) |
+-------------------+-----------------+----------------+----------------+
| AIC               | ```             | ```            | ```            |
|                   | -885.7352       | -887.8655      | -883.9013      |
|                   | ```             | ```            | ```            |
+-------------------+-----------------+----------------+----------------+
| Q-Statistic       | ```             | ```            | ```            |
|                   | 4.6056          | 4.7308         | 4.7989         |
|                   | ```             | ```            | ```            |
+-------------------+-----------------+----------------+----------------+
| p-value of Q-Stat | ```             | ```            | ```            |
|                   | 0.708           | 0.8571         | 0.7788         |
|                   | ```             | ```            | ```            |
+-------------------+-----------------+----------------+----------------+

: Table 1

# Forecasting

We will forecast for :

```{r}
length(weekly_return_hb)
```

`74` weeks ahead.

```{r}
ar1_forecast <- forecast(ar1, h = 74)
plot(ar1_forecast)
```

```{r}
ts.plot(ar1_forecast$mean, weekly_return_hb, lty=c(1,2))
```

The forecast seems to not have any significant variation

**adding higher order arma models**

```{r}
arma192 <- arima(weekly_return_est$`BTC-USD.Close`,order=c(19,0,2))
summary(arma192)
```

```{r}
coeftest(arma192)
```

```{r}
checkresiduals(arma192)
```

Stable residuals with stationarity

**Forecasts:**

```{r}
arma192_forecast<-forecast(arma192, h=74)
plot(arma192_forecast)
```

```{r}
ts.plot(arma192_forecast$mean, weekly_return_hb$`BTC-USD.Close`, lty=c(1,2))
```

**ARMA(2,19)**

```{r}
arma219 <- arima(weekly_return_est$`BTC-USD.Close`,order=c(2,0,19))
summary(arma219)
```

```{r}
coeftest(arma219)
```

```{r}
checkresiduals(arma219)
```

```{r}
arma219_forecast<-forecast(arma219,h=74)
plot(arma219_forecast)
```

```{r}
ts.plot(arma219_forecast$mean, weekly_return_hb$`BTC-USD.Close`, lty=c(1,2))
```

```{r}
weekly_return_hb_ts <- as.ts(weekly_return_hb,start=2927)
errors_ar1 <- weekly_return_hb_ts - ar1_forecast$mean
errors_arma192 <- weekly_return_hb_ts - arma192_forecast$mean
errors_arma219 <- weekly_return_hb_ts - arma219_forecast$mean
```

```{r}
mspear1 <- mean(errors_ar1^2)
mspearma192 <- mean(errors_arma192^2)
mspearma219 <- mean(errors_arma219^2)
```

### Paired F-test

#### AR1 ARMA192

```{r}
f_ararma192 <- ((mspear1 - mspearma192) / (74 - 1)) / (mspearma192 / (74-21))
p_value_f <- 1 - pf(f_ararma192, df1 = 74 - 1, df2 = 74-21)
p_value_f
```

The p-value of 1 suggests that there is not enough evidence to reject the null hypothesis that the two models have the same Mean Squared Prediction Error (MSPE). A p-value of 1 indicates that the difference in MSPE between the two models is likely due to random chance, rather than a true difference in forecast accuracy.

There might also be a chance of overfit.

#### AR1 ARMA(2,19)

```{r}
f_ararma219 <- ((mspear1 - mspearma219) / (74 - 1)) / (mspearma219 / (74-21))
p_value_f <- 1 - pf(f_ararma219, df1 = 74 - 1, df2 = 74-21)
p_value_f
```

Same p-value

#### ARMA(19,2) ARMA(2,19)

```{r}
f_armaarma <- ((mspearma192 - mspearma219) / (74 - 21)) / (mspearma219 / (74-21))
p_value_f <- 1 - pf(f_armaarma, df1 = 74 - 21, df2 = 74-21)
p_value_f
```

Same p-value

### DM Test

```{r}
dm_test_1<- dm.test(errors_ar1^2, errors_arma192^2, h = 74)
dm_test_1
```

```{r}
dm_test_2<- dm.test(errors_ar1^2, errors_arma219^2, h = 74)
dm_test_2
```

```{r}
dm_test_3<- dm.test(errors_arma192^2, errors_arma219^2, h = 74)
dm_test_3
```

#### Absolute loss function

Since Absolute loss penalizes large errors linearly, whereas quadratic loss penalizes them quadratically, we can have a better differentiation wrt larger errors.

```{r}
dm_test_ar1_arma192 <- dm.test(abs(errors_ar1), abs(errors_arma192), alternative = "two.sided", h = 1)
dm_test_ar1_arma219 <- dm.test(abs(errors_ar1), abs(errors_arma219), alternative = "two.sided", h = 1)
dm_test_arma192_arma219 <- dm.test(abs(errors_arma192), abs(errors_arma219), alternative = "two.sided", h = 1)

# Print DM test results
cat("DM test AR(1) vs ARMA(19,2):", "statistic =", dm_test_ar1_arma192$statistic, "p-value =", dm_test_ar1_arma192$p.value, "\n")
cat("DM test AR(1) vs ARMA(2,19):", "statistic =", dm_test_ar1_arma219$statistic, "p-value =", dm_test_ar1_arma219$p.value, "\n")
cat("DM test ARMA(19,2) vs ARMA(2,19):", "statistic =", dm_test_arma192_arma219$statistic, "p-value =", dm_test_arma192_arma219$p.value, "\n")

```

### Table 2

+--------------------------+----------------+--------------+
|                          | Test Statistic | p-value      |
+==========================+================+==============+
| AR(1) vs ARMA(19,2)      |                |              |
+--------------------------+----------------+--------------+
| F-test                   | ```            | 1            |
|                          | -9.87831e-05   |              |
|                          | ```            |              |
+--------------------------+----------------+--------------+
| DM-test (Quadratic Loss) | ```            | ```          |
|                          | 0.5157         | 0.6076       |
|                          | ```            | ```          |
+--------------------------+----------------+--------------+
| DM-test (Abs Loss)       | -0.0047        | 0.99621      |
+--------------------------+----------------+--------------+
| AR(1) vs ARMA(2,19)      |                |              |
+--------------------------+----------------+--------------+
| F-test                   | ```            | 1            |
|                          | 0.00620222     |              |
|                          | ```            |              |
+--------------------------+----------------+--------------+
| DM-test (Quadratic Loss) | ```            | 1            |
|                          | 0              |              |
|                          | ```            |              |
+--------------------------+----------------+--------------+
| DM-test (Abs Loss)       | ```            | ```          |
|                          | 0.2154764      | 0.8299968    |
|                          | ```            | ```          |
+--------------------------+----------------+--------------+
| AR(19,2) vs ARMA(2,19)   |                |              |
+--------------------------+----------------+--------------+
| F-test                   | ```            | 1            |
|                          | 0.008679921    |              |
|                          | ```            |              |
+--------------------------+----------------+--------------+
| DM-test (Quadratic Loss) | ```            | ```          |
|                          | 0.58104        |  0.563       |
|                          | ```            | ```          |
+--------------------------+----------------+--------------+
| DM-test (Abs Loss)       | ```            | ```          |
|                          | 0.3078686      | 0.7590591    |
|                          | ```            | ```          |
+--------------------------+----------------+--------------+

```{r}
mean_forecast_ar1 <- mean(ar1_forecast$mean)
mean_forecast_arma192 <- mean(arma192_forecast$mean)
mean_forecast_arma219 <- mean(arma219_forecast$mean)

squared_diff_ar1 <- (ar1_forecast$mean - mean_forecast_ar1)^2
forecast_variance_ar1 <- mean(squared_diff_ar1)

squared_diff_arma192 <- (arma192_forecast$mean - mean_forecast_arma192)^2
forecast_variance_arma192 <- mean(squared_diff_arma192)

squared_diff_arma219 <- (arma219_forecast$mean - mean_forecast_arma219)^2
forecast_variance_arma219 <- mean(squared_diff_arma219)

cat("Forecast Variance AR(1):", forecast_variance_ar1, "\n")
cat("Forecast Variance ARMA(19,2):", forecast_variance_arma192, "\n")
cat("Forecast Variance ARMA(2,19):", forecast_variance_arma219, "\n")

```

AR(1) shows least variance but without trend.

Changing the estimation and holdback

```{r}
btc_week_est <- btc_week_df[index(btc_week_df) >= as.Date("2014-09-21") & 
                                index(btc_week_df) <= as.Date("2023-07-21"), ]

btc_week_hb <- btc_week_df[index(btc_week_df) >= as.Date("2023-07-22") & 
                              index(btc_week_df) <= as.Date("2024-02-23"), ]
```

80-20

```{r}
weekly_return_est <- weekly_return[index(weekly_return) >= as.Date("2014-09-21") & 
                                index(weekly_return) <= as.Date("2023-07-21"), ]
weekly_return_hb <- weekly_return[index(weekly_return) >= as.Date("2023-07-22") & 
                                index(weekly_return) <= as.Date("2024-02-23"), ]
```

\

```{r}
ar1<-arima(weekly_return_est$`BTC-USD.Close`,order=c(1,0,0))
print(summary(ar1))

arma192 <- arima(weekly_return_est$`BTC-USD.Close`,order=c(19,0,2))
print(summary(arma192))

arma219 <- arima(weekly_return_est$`BTC-USD.Close`,order=c(2,0,19))
print(summary(arma219))
```

```{r}
arma219_forecast<-forecast(arma219,h=31)
plot(arma219_forecast)
```

```{r}
arma192_forecast<-forecast(arma192,h=31)
plot(arma192_forecast)
```

```{r}
ar1_forecast<-forecast(ar1,h=31)
plot(ar1_forecast)
```

```{r}
ts.plot(arma219_forecast$mean, weekly_return_hb, lty=c(1,2))
```

```{r}
ts.plot(ar1_forecast$mean, weekly_return_hb, lty=c(1,2))
```

```{r}
ts.plot(arma192_forecast$mean, weekly_return_hb, lty=c(1,2))
```

```{r}
weekly_return_hb_ts <- as.ts(weekly_return_hb,start=3228)
errors_ar1 <- weekly_return_hb_ts - ar1_forecast$mean
errors_arma192 <- weekly_return_hb_ts - arma192_forecast$mean
errors_arma219 <- weekly_return_hb_ts - arma219_forecast$mean
mspear1 <- mean(errors_ar1^2)
mspearma192 <- mean(errors_arma192^2)
mspearma219 <- mean(errors_arma219^2)
print(mspear1)
print(mspearma192)
print(mspearma219)
```

The error has been reduced but the is still similar (lack of) trend.

```{r}
mean_forecast_ar1 <- mean(ar1_forecast$mean)
mean_forecast_arma192 <- mean(arma192_forecast$mean)
mean_forecast_arma219 <- mean(arma219_forecast$mean)

squared_diff_ar1 <- (ar1_forecast$mean - mean_forecast_ar1)^2
forecast_variance_ar1 <- mean(squared_diff_ar1)

squared_diff_arma192 <- (arma192_forecast$mean - mean_forecast_arma192)^2
forecast_variance_arma192 <- mean(squared_diff_arma192)

squared_diff_arma219 <- (arma219_forecast$mean - mean_forecast_arma219)^2
forecast_variance_arma219 <- mean(squared_diff_arma219)

cat("Forecast Variance AR(1):", forecast_variance_ar1, "\n")
cat("Forecast Variance ARMA(19,2):", forecast_variance_arma192, "\n")
cat("Forecast Variance ARMA(2,19):", forecast_variance_arma219, "\n")

```

AR1 returns minimum forecast variance, but doesnt show any trend.

Despite inaccuracy due to seasonality of data we do have consistent results.
